\Large Characters' psychological arrows of time drive temporal asymmetries in observers' inferrences about past and future narrative events

### 2023-10-09 paper/old.tex
checking in changes from overleaf

\title{\Large The psychological arrow of time drives temporal asymmetries in inferring unobserved past and future events}
\author[1]{Xinming Xu}
\author[2]{Ziyan Zhu}
\author[1, $\star$]{Jeremy R. Manning}
\affil[1]{Dartmouth College, Hanover, NH, USA}
\affil[2]{Peking University, Beijing, China}
\affil[$\star$]{Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract} {
  How much can we infer about the past and future, given our knowledge of the present?  Unlike temporally symmetric inferences about simple sequences, inferences about our own lives are asymmetric: we are better able to infer the past than the future, since we remember our past but not our future (i.e., the psychological arrow of time).  What happens when both the past and future are unobserved, as when we make inferences about \textit{other} people's lives?  We had participants view segments of a character-driven television drama.  They wrote out what would happen just before or after each just-watched segment.  Participants were better at inferring past (versus future) events.  This asymmetry was driven by participants’ reliance on characters’ conversational references in the narrative, which tended to favor the past.  Our work reveals a temporal asymmetry in how observations of other people’s behaviors can inform us about the past and future.

\textbf{Keywords: arrow of time, prediction, retrodiction, narrative, conversation}}

@@

\section*{Introduction}
What we experience in the current moment tells us about \textit{now}-- but what does it tell us about the past or future?  And does the current moment tell us, as human observers, \textit{more} about the past or about the future?  One way of examining these questions is to consider highly simplified scenarios that are artificially constructed in the laboratory \citep[e.g., ][]{MaheEtal22}.  At one extreme, for deterministic sequences with \textit{known} rules, knowing the current state provides the observer with sufficient information to exactly reconstruct the entire past and future history of the stimulus. At another extreme, for purely random sequences, observing the current state provides no information about the past \textit{or} future.

Sequences generated by stochastic processes fall somewhere between these two extremes. For Markov processes, where each state is solely dependent on the immediately preceding state, Shannon entropy may be used to quantify the uncertainty of the past and future states, given the present state. \cite{Cove94} showed that, for any stationary process (i.e., processes in equilibrium), Markov or otherwise, the present state provides equal information (i.e., mutual information) about past and future states~\citep[also see][]{BialEtal01, ElliEtal09}. Further, there is some evidence that humans are similarly adept at inferring the most likely previous and next items in sequences governed by stochastic Markov processes~\citep{JonePash07}.

Deterministic, random, and probabilistic sequences (in equilibrium) are all symmetric: the present state of these sequences is equally informative about past versus future states.  In contrast, our subjective experience in everyday life is that we know more about our own past than our future~\citep[e.g.,][]{Horw87}.  We have memories of our past that we carry with us into the present moment, but we do not have memories of our yet-to-be-experienced future.  This temporal asymmetry imposes an ``arrow of time'' on our subjective experience, known as the \textit{psychological arrow of time}~\citep[e.g.,][]{Hawk85}.

@@

Although the psychological arrow of time implies that we should be better able to infer our past than our future, how generally does this temporal asymmetry hold?  And does the asymmetry hold only for our own experiences (due to our memories), or is the asymmetry a general property of any real-life event sequence?  In real-world situations (and narratives) where we are \textit{equally} ignorant of the past and future, as for \textit{other} people's lives where we lack memories of the relevant past, are our inferences about the past and future symmetric or asymmetric?  For example, imagine that you are meeting a stranger for the first time.  At the moment of your meeting, you lack both memories of their past and knowledge about what they might do in the future.  After your first encounter with the stranger, would you be able to more accurately or easily form inferences about what had happened in their past (\textit{retrodiction}) or what will happen in their future (\textit{prediction}; Fig.~\ref{fig:intro1})?  Or suppose you started watching a movie partway through.  Again, you would enter the moment of watching without memories of prior parts of the movie. Given your observations in the present, would your guesses about what had happened before you started watching be more (or less) accurate than your guesses about what will happen next?  In general, when the past and future are \textit{both} unobserved, are we better at inferring the past or the future in real-world settings?  Narrative stimuli, such as stories and movies, can provide a useful testbed for exploring several of these questions.

Although narratives are unlikely to be confused with one's own experiences, narratives mirror some of the structure of real-world experiences.  Character behaviors and interactions are often designed in a way that helps the audience connect with or relate to the characters.  Events in narratives also unfold in ways that are intended to build rapport or engagement with the audience.  This might be accomplished by having events follow a believable structure that is reminiscent of real-world experiences, or by designing the audience's experiences in ways that communicate clear ``rules'' or ``features'' that help to immerse the audience in the narrative's universe.  The characters in a realistic narrative can also be written to behave in ways reminiscent of real-world people.  These same aspects of narratives that authors use to drive engagement with events and characters can lead narratives to replicate some core aspects of real-world experiences that are typically lost or overlooked in traditional sequence learning paradigms.  Narratives can drive the audience to build situation models~\citep{RadvCope06, ZwaaRadv98} of the narrative's universe, or to form a theory of mind of and make predictions about the characters~\citep{TamiThor18, KostSaxe13}.  Events in narratives may unfold in a consistent or logical way, but they also exhibit complex and meaningful interactions across events reminiscent of real-world experiences (but not necessarily the simple sequences traditionally used in the statistical learning literature).

One key difference between simple artificial sequences and more naturalistic (real or narrative) sequences is that naturalistic sequences often incorporate other people.  Despite the past and future being equally unknown to \textit{the observer} prior to the current moment, other people, and realistic characters in narratives, have their own psychological arrows of time. Specifically, they have memories of their own pasts. Other people's asymmetric knowledge about their \textit{own} pasts and futures might affect their behaviors (e.g., conversations). In turn, this might provide time-asymmetric clues that favor the past~\citep[e.g., other people might talk more about their own pasts than their futures; ][]{DemiEtal18}. If observers leverage these clues from other people's asymmetric knowledge, then observers should also be better at inferring the past (versus the future) of other people's lives.  Alternatively, if inferences about other people's lives are more like inferences about artificial statistical sequences~\citep[e.g., perhaps solely relying on statistical regularities like event schemas, scripts, or situation models][]{RadvCope06, ZwaaRadv98, BoweEtal79, RangRitc12}, then the accuracy of inferences about the past and the future of others' lives should be approximately equal.

We designed a naturalistic paradigm for exposing participants to scenarios where the past and future were equally unobserved.  We asked our participants to watch a series of movie segments drawn from a character-driven dramatic television show.  Across the conditions and trials in the experiment, participants made free-form text responses to either retrodict what had happened in the previous segment, predict what would happen in the next segment, or recall what happened in the just-watched segment.  We used manual annotations and sentence-level natural language processing models to characterize participants' responses.  To foreshadow our results, we found that participants were overall better at retrodicting the past than predicting the future. This appeared to be driven by two main factors.  First, characters more often referred to past events than future (e.g., planned) events, and this influenced participants' responses. Second, associations and dependencies between temporally adjacent events enabled participants to form estimates about nearby events (e.g., to a just-watched scene or a past or future event referenced in an observed conversation).  Taken together, our work reveals a temporal asymmetry in how observations of other humans’ behaviors inform us about the past versus the future.

\section*{Results}
Participants in our study ($n = 36$) watched segments from two storylines, drawn from the CBS television show \textit{Why Women Kill}.  Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table \stimDescription).  We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched (Fig.~\ref{fig:method}, \textit{Task design}). We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the \textit{target segment} for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments they had seen prior to making a response (see \textit{Methods}).

@@

We asked participants to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}).  To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment.  To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment.  We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions.  To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment.  We labeled these responses according to which other segments participants had watched prior to the just-watched target.  Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2; Fig.~\ref{fig:method}).  Whereas retrodiction and prediction responses reflect what participants \textit{estimate} they would remember after watching the (inferred) target segment, recall responses provide a benchmark for comparison by measuring what they \textit{actually} remember about the target segment.

@@

We used two general approaches to assess the quality of participants' responses (see \textit{Methods}, Fig.~\ref{fig:result1}A).  One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses.  We identified a total of 117 unique events reflected across the 22 video segments (range: 3--9 per segment; see \textit{Methods}, Table \stimDescription).  We assigned one ``point'' to each of these video events.  We also identified 23 additional events in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events.  We assigned 0.5 point to each of these additional events.  This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response.  Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space.  This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events.  To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses.  Following \cite{HeusEtal21}, we defined the \textit{precision} of each participants' retrodictions or predictions about a target segment as the median cosine similarities between the embeddings of (a) the participant's retrodiction or prediction response for the target segment and (b) each \textit{other} participant's recalls of the same segment.  In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered.  We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants.  In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a target segment and all other participants' responses (of the same type) to the same segment.  We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Methods}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction).  Across these three types of responses, participants have access to increasing amounts of information about the target segment.  Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals.  Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B).  These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D).   Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and were more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses.  We hoped that these analyses might provide insights into our central question about whether inferences about the past and future are equally accurate.  Across both uncued and character-cued responses (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B).  However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52).  We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C).  Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11).  We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09).  Taken together, these results suggest that participants are generally better at making retrodictions than predictions.  We also verified that this was not solely a consequence of how participants' memory performance might have been affected by watching different segments (or making different responses to other segments) across conditions by comparing recall responses in the retrodiction-matched recall (\textit{re(R)}) and prediction-matched recall (\textit{re(P)}) conditions.  Recall performance was similar in both conditions (target event hit rate: OR = 1.12, $Z = 1.07$, $p = 0.29$, CI: 0.91 to 1.39; convergence: $b = 0.03$, $t(19.3) = 1.89$, $p = 0.07$, CI: 0.00 to 0.07).

@@

The above analyses were focused solely on the target segment (i.e., retrodiction of segment $n$ after watching segments $(n+1)... 11$, or prediction of segment $n$ after watching segments $1 ...(n-1)$).  We wondered whether participants' responses might also contain longer-range information about preceding or proceeding events.  In order to carry out this analysis properly, we reasoned that participants might reference past or future events that were \textit{implied} to have occurred offscreen, but not explicitly shown onscreen.  For example, a character in location A during one scene might appear in location B during the immediately following scene.  Although it wasn't shown onscreen, we can infer that the character traveled between locations A and B sometime between the time intervals separating the scenes~\citep{Bord08}.  In all, we manually identified a set of 74 \textit{implicit} offscreen events that were implied to have occurred given what was (explicitly) depicted onscreen (Fig.~\ref{fig:result2}A), plus one additional partial event and one additional summary event.  We defined the just-watched segment as having a \textit{lag} of 0.  We assigned the target segment of a participant's retrodiction or prediction (i.e., the immediately preceding or proceeding segment) a lag of -1 or +1, respectively.  The segment following the next was assigned a lag of 2, and so on.  We tagged offscreen events using half steps.  For example, an offscreen event that occurred after the prior segment but before the just-watched segment would be assigned a lag of -0.5.

@@

@@

An additional ambiguous case arose in one participant's responses pertaining to segment 12, storyline 2, whereby the participant correctly identified an onscreen event that had not been included in our original annotations.  To account for this participant's response, we retroactively added that event to our annotations of that segment.  We also identified and counted unmatched events in participants' responses (i.e., events that did not match any annotated events).  Cases where the two coders' independent scoring disagreed were resolved through discussions between the two coders.

>>>>

\title{\Large Characters' psychological arrows of time drive temporal asymmetries in observers' inferrences about past and future narrative events}
\author[1]{Xinming Xu}
\author[2]{Ziyan Zhu}
\author[1, $\star$]{Jeremy R. Manning}
\affil[1]{Dartmouth College, Hanover, NH, USA}
\affil[2]{Peking University, Beijing, China}
\affil[$\star$]{Address correspondence to jeremy.r.manning@dartmouth.edu}

@@

\begin{abstract} {\footnotesize
  How much can we infer about the past and future, given our current knowledge and observations in the present?  Inferences about our own lives are time-asymmetric: we are better able to infer the past than the future, since we remember our past but not our future.  The asymmetry in our memories is known as the \textit{psychological arrow of time}.  But when both the past and future are unexperienced and unobserved, for example when we make inferences about \textit{other} people's lives, are our inferences about the past and future time-symmetric or asymmetric?  To study these questions, we had participants view segments of a character-driven television drama.  They used free-form responses to guess at what would happen just before or just after each just-watched segment.  We found that participants' inferences were time-asymmetric, in that they inferred past events more accurately than future events after controlling for their exposure to past and future events in the narrative.  This asymmetry was driven by \textit{characters'} biases in conversational references to their own pasts.  Our work reveals a temporal asymmetry in how observations of other peoples’ behaviors can inform us about the past and future.

\textbf{Keywords: episodic memory, prediction, retrodiction, narratives, conversation}}

@@

\section*{Introduction}

What we experience in the current moment tells us about \textit{now}-- but what does it tell us about the past or future?  One way of examining this question is to consider highly simplified scenarios that are artificially constructed in the laboratory.  At one extreme, for fully observable deterministic sequences, observing any part of the sequence provides sufficient information to reconstruct the entire past and future history of the stimulus.  For example, if we observe the partial sequence ``5, 6, 7, 8, 9, 10,'' we might reasonably guess that the just-preceding number was 4, and the just-proceeding number will be 11.  If we were given additional information through further observations, or by being told about the rules for generating the sequence, our confidence would increase.  Eventually, with more observations or more knowledge about the underlying rules, our uncertainty might approach zero.  At another extreme, for purely random sequences, observing the current state provides no information about the past or future.

The statistical learning literature~\citep{SchaTurk15, Hass17, MaheEtal22} has tended to focus on manufactured scenarios between these two extremes, whereby participants experience structured (but not fully deterministic) sequences.  Often the sequences encountered in this domain are Markov processes, whereby (in the simplest framing, as in first-order Markov processes) each element of the sequence is solely dependent on the immediately preceding element, plus some additional noise~\citep{McNeEtal06, CuniEtal09, FurlEtal11, DaikEtal14, DaikEtal15, KoelEtal16, JonePash07, TummEtal16}.  In the typical setup, participants are instructed to adjust their behaviors or to form explicit predictions about what will come next, given their observations up to that point.  In some studies, participants are also asked to generate inferences about the unobserved past (i.e., what came \textit{earlier}) in a sequence, prior to their first observation~\citep{JonePash07}.

Markov sequences are, by definition, symmetric in time: after learning the ``rules'' (i.e., transition probabilities) that govern how the sequences are generated, observing any part of the sequence provides equal information about its past and future states~\citep{Cove94, BialEtal01, ElliEtal09}.  By contrast, real-world experiences are often asymmetric, since our memories provide us with information about our past experiences but not our yet-to-be-experienced future.  This memory-driven asymmetry in our knowledge is known as the \textit{psychological arrow of time}~\citep[e.g.,][]{Hawk85}.  Although the notion of the psychological arrow of time helps to formalize why we know more about our \textit{own} past than our future, how generally does this hold?  In real-world situations where we are \textit{equally} ignorant of the past and future, as for \textit{other} people's lives where we lack memories of the relevant past, is our knowledge about the past and future symmetric or asymmetric (Fig.~\ref{fig:intro1})?  For example, imagine that you are meeting a stranger for the first time.  Would you be able to more accurately or easily form inferences about what had happened in their past versus their future~\citep{TamiThor18, KostSaxe13}?  Or suppose you started watching a movie partway through.  Again, given your observations in the present, would your guesses about what had happened before you started watching be more (or less) accurate than your guesses about what will happen next?  In general, when the past and future are \textit{both} unobserved, are we better at inferring the past or the future in real-world settings?

@@

If we hope to connect prior work on statistical learning tasks that employ sequences generated using relatively simple rules with the sorts of inferences we make in everyday life, several considerations apply.  One set of considerations concerns how similar the underlying neurocognitive processes are when we encounter simple sequences in the laboratory versus more complex sequences in our real-world experiences.  In other words, if our day-to-day experiences violate the Markov assumption by displaying interactions between temporally separated occurrences or events, does this change anything meaningful in terms of how our brains process, respond to, leverage, or remember those experiences?  Another set of considerations concerns how similar the experiences themselves are.  For example, to what extent are real-world experiences deterministic versus stochastic?  To what extent are real-world experiences structured versus random?  And to what extent do real-world experiences satisfy the Markov assumption?  Narrative stimuli, such as stories and movies, can provide a useful testbed for exploring several of these questions.

Although narratives are unlikely to be confused with one's own experiences, narratives are most compelling when they mirror some of the structure of real-world experiences.  Character behaviors and interactions are often designed in a way that helps the audience connect with or relate to the characters.  Events in narratives also unfold in ways that are intended to build rapport or engagement with the audience.  This might be accomplished by having events follow a believable structure that is reminiscent of real-world experiences, or by designing the audience's experiences in ways that communicate clear ``rules'' or ``features'' that help to immerse the audience in the narrative's universe.  These same aspects of narratives that authors use to drive engagement with characters and events can lead narratives to replicate some core aspects of real-world experiences that are typically lost or overlooked in traditional sequence learning paradigms.  Narratives can drive the audience to form a theory of mind of the characters~\citep{TamiThor18}, or to build situation models~\citep{RangRitc12, BoweEtal79} of the narrative's universe.  Events in narratives may unfold in a consistent or logical way, but they also often incorporate stochastic elements and events that violate the Markov assumption, such as when the consequences of an action or event early on in the narrative are realized much later on.

If inferences formed for real-world events and narratives are like inferences for the Markov sequences often studied in the statistical learning literature, people should be equally adept at inferring the past and future.  Alternatively, perhaps important differences between naturalistic versus Markov sequences might lead to an advantage for the past or future.  In the above examples of meeting a stranger or starting a movie partway through, perhaps the \textit{stranger's} (or characters') asymmetric knowledge about their own pasts might influence their observable behaviors.  Despite the past and future being equally unknown to \textit{you} prior to the current moment, other people's asymmetric knowledge about their own observed pasts might give you an advantage at inferring the unobserved past versus future~\citep{PillEtal15}.  Similarly, if a conversation you joined partway through was focused on planning a future event, that might advantage your inferences of the future (versus the past).  In general, what sorts of behaviors might reflect other individuals' asymmetric knowledge about their past, or their plans for the future?  Are there particular types of information that people are especially sensitive to, or that are particularly informative or useful?   Could we quantify or characterize those factors in a systematic way?  Or might they be too subtle or nuanced to formally define?

At a high level, these questions all relate to how people form inferences about the past and future, given their observations.  But these questions also often relate to issues of temporal symmetry-- i.e., whether inferring the past is \textit{easier} or \textit{harder} than inferring the future.  Presumably, the overall amount of information present will determine something about how \textit{much} may be inferred before or after the present.  In other words, a tiny amount of information, or an unreliable observation, and so on, will provide less insight into other moments than a large amount of reliable information.  In a sense, this is trivial: more information is ``better'' if our goal is to form inferences.  However, \textit{controlling} for the amount of information, can we learn anything meaningful about inherent biases in either our memory systems (including related systems for forming inferences) or in the sort of information we are likely to encounter?  Studying these questions often entails comparing inferences about the past versus the future by characterizing any apparent symmetries or asymmetries.

We designed a naturalistic paradigm for exposing participants to scenarios where the past and future were equally unobserved.  We asked our participants to watch a series of movie segments drawn from a character-driven dramatic television show.  Across the conditions and trials in the experiment, we carefully controlled for how much of the past or future of the narrative the participants had experienced prior to viewing the target segment.  We then asked the participants to guess about what might have happened before the target segment (\textit{retrodiction}), what might happen after the target segment (\textit{prediction}), or to recount what they had observed in the target segment itself (\textit{recall}).  We used manual annotations and sentence-level natural language processing models to characterize participants' responses.

\section*{Results}
Participants in our study ($n = 36$) watched segments from two storylines, drawn from the CBS television show, \textit{Why Women Kill}.  Each storyline comprised 11 segments (mean duration: 2.05~min; range: 0.97--3.87~min, Table \stimDescription).  We asked participants to use free-form (typed) text responses to retrodict what had happened prior to a just-watched segment, predict what would happen next, or recall what they had just watched. We referred to the to-be-retrodicted, to-be-predicted, or to-be-recalled segment as the target segment for each response. We systematically varied whether participants watched the segments in forward or reverse chronological order, and how many segments preceded or proceeded the target segment they had seen prior to making a response (Fig.~\ref{fig:method}, \textit{Task design}).

@@

We asked participants to generate four types of responses after watching each video segment: uncued responses, character-cued responses, updated responses, and recalls (Fig.~\ref{fig:method}, \textit{Data overview}).  To generate \textit{uncued} responses, we asked participants to either retrodict (uncued retrodiction; \textit{u-R}) what happened shortly before or predict (uncued prediction; \textit{u-P}) what happened shortly after the just-watched segment.  To generate \textit{character-cued} responses, we asked participants to retrodict (character-cued retrodiction; \textit{c-R}) or predict (character-cued prediction; \textit{c-P}) what came before or after the just-watched segment, but we provided additional information to the participant about which character(s) would be present in the target (to-be-retrodicted or to-be-predicted) segment.  We hypothesized that character-cued responses should be more accurate than uncued responses, to the extent that participants incorporate the character information we provided to them into their retrodictions and predictions.  To generate updated responses, we asked participants to watch an additional segment that came just prior to or just after the target segment, and then to update their retrodiction (\textit{c-RP}) or prediction (\textit{c-PR}) about the target segment. Results on updated responses are not reported in this paper. Finally, we also asked participants to \textit{recall} what happened in the just-watched segment.  We labeled these responses according to which other segments participants had watched prior to the just-watched target.  Retrodiction-matched recall (\textit{re(R)}) responses were made during the retrodiction sequences (B1 and B2; Fig.~\ref{fig:method}), whereas prediction-matched recall (\textit{re(P)}) responses were made during the prediction sequences (A1 and A2).  Participants' recalls provided us with a benchmark for examining information about the participants' experiences and memories, without asking the participants to explicitly speculate about the unobserved.

@@

We used two general approaches to assess the quality of participants' responses (see \textit{Methods}, Fig.~\ref{fig:result1}A).  One approach entailed manually annotating events in the video and counting the number of matched events in participants' responses.  We identified a total of 117 unique events reflected across the 22 video segments (range: 3--9 per segment; see \textit{Methods}, Table \stimDescription).  We assigned one ``point'' to each of these video events.  We also identified 23 additional events in participants' responses that were either summaries of several events or that were partial matches to the manually identified video events.  We assigned 0.5 point to each of these additional events.  This point system enabled us to compute the numbers and proportions (\textit{hit rates}) of correctly retrodicted, predicted, and recalled events contained in each response.  Our second approach entailed using a natural language processing model~\citep{CerEtal18} to embed annotations and responses in a 512-dimensional feature space.  This approach was designed to capture conceptual overlap between responses that were not necessarily tied to specific events.  To quantify this conceptual overlap, we computed the similarities between the embeddings of different sets of responses.  Following \cite{HeusEtal21}, we defined the \textit{precision} of each participants' retrodictions or predictions about a given segment as the median cosine similarities between the embeddings of (a) the participant's retrodiction or prediction response for the given segment and (b) each \textit{other} participant's recalls of the same segment.  In other words, precision is designed to measure the extent to which retrodictions and predictions captured the conceptual content that (other) participants remembered.  We also developed a related measure, which we call \textit{convergence}, to characterize response similarities across participants.  In particular, we defined convergence as the mean cosine similarity between the embeddings of a participant's responses to a given target segment and all other participants' responses (of the same type) to the same segment.  We analyzed the data using generalized linear mixed models, with participant and stimulus (e.g., target segment) identities as crossed random effects (see \textit{Methods}).

First we sought to validate a main effect of response type (i.e., uncued responses, character-cued responses, and recalls), irrespective of the temporal direction (retrodiction versus prediction).  Across these three types of responses, participants have access to increasing amounts of information about the target segment.  Therefore, across these response types, we hypothesized that participants' responses should become both more accurate and more convergent across individuals.  Consistent with this hypothesis, participants' character-cued retrodictions and predictions were associated with higher target event hit rates than uncued retrodictions and predictions (odds ratio (OR): 2.65, $Z = 4.24$, $p < 0.001$, 95\% confidence interval (CI): 1.69 to 4.16; Fig.~\ref{fig:result1}B).  These character-cued responses were also more precise ($b = 0.13$, $t(18.1) = 9.43$, $p < 0.001$, CI: 0.10 to 0.16; Fig.~\ref{fig:result1}C) and convergent across individuals ($b = 0.11$, $t(18.6) = 6.21$, $p < 0.001$, CI: 0.07 to 0.15; Fig.~\ref{fig:result1}D).   Relative to character-cued responses, participants' recalls showed higher target event hit rates (OR = 21.83, $Z = 10.61$, $p < 0.001$, CI: 12.35 to 38.59) and more convergence across individuals ($b = 0.20$, $t(19.4) = 9.10$, $p < 0.001$, CI: 0.16 to 0.25). These results are consistent with the common-sense notion that access to more information about a target segment yields better performance (i.e., higher hit rates, precision, and convergence across individuals).

Next we carried out a series of analyses specifically aimed at characterizing temporal direction effects--- i.e, the relative quality of retrodictions versus predictions across different types of responses.  We hoped that these analyses might provide insights into our central question about whether the present is equally informative about the past and future.  Across both uncued and character-cued responses (Fig.~\ref{fig:method}), retrodictions had numerically higher hit rates than predictions (Fig.~\ref{fig:result1}B).  However, these differences were only statistically reliable for character-cued responses (uncued responses: OR = 1.17, $Z = 0.35$, $p = 0.73$, CI: 0.47 to 2.92; character-cued responses: OR = 1.93, $Z = 2.15$, $p = 0.03$, CI: 1.06 to 3.52).  We observed a similar pattern of results for the precisions of participants' responses (Fig.~\ref{fig:result1}C).  Specifically, their responses tended to be numerically more precise for retrodictions versus predictions, but the differences were only statistically reliable for character-cued responses (uncued responses: $b = 0.03$, $t(20.9) = 1.09$, $p = 0.29$, CI: -0.03 to 0.10; character-cued responses: $b = 0.06$, $t(20.8) = 3.01$, $p = 0.007$, CI: 0.02 to 0.11).  We also consistently observed numerically higher convergence across participants for retrodictions versus predictions (Fig.~\ref{fig:result1}D), but neither of these differences were statistically reliable (uncued responses: $b = 0.03$, $t(17.9) = 0.75$, $p = 0.46$, CI: -0.05 to 0.11; character-cued responses: $b = 0.04$, $t(17.4) = 1.46$, $p = 0.16$, CI: -0.02 to 0.09).  Because the retrodiction versus prediction performance differences we observed were only statistically reliable when participants were cued with the target segments' characters, this suggests that information about the unobserved past versus the unobserved future may differently affect retrodictions versus predictions.  Taken together, these results suggest that participants are generally better at making retrodictions than predictions.  We also verified that this was not solely a consequence of how participants' memory performance might have been affected by watching different segments (or making different responses to other segments) across conditions by comparing recall responses in the retrodiction-matched recall (\textit{re(R)}) and prediction-matched recall (\textit{re(P)}) conditions.  Recall performance was similar in both conditions (target event hit rate: OR = 1.12, $Z = 1.07$, $p = 0.29$, CI: 0.91 to 1.39; convergence: $b = 0.03$, $t(19.3) = 1.89$, $p = 0.07$, CI: 0.00 to 0.07).

@@

The above analyses were focused solely on the target segment (i.e., retrodiction of segment $i - 1$ after watching segments $i...11$, or prediction of segment $i + 1$ after watching segments $1...i$).  We wondered whether participants' responses might also contain longer-range information about preceding or proceeding events.  In order to carry out this analysis properly, we reasoned that participants might reference past or future events that were \textit{implied} to have occurred offscreen, but not explicitly shown onscreen.  For example, a character in location A during one scene might appear in location B during the immediately following scene.  Although it wasn't shown onscreen, we can infer that the character traveled between locations A and B sometime between the time intervals separating the scenes~\citep{Bord08}.  In all, we manually identified a set of 74 \textit{implicit} offscreen events that were implied to have occurred given what was (explicitly) depicted onscreen (Fig.~\ref{fig:result2}A), plus one additional partial event and one additional summary event.  We defined the just-watched segment as having a \textit{lag} of 0.  We assigned the target segment of a participant's retrodiction or prediction (i.e., the immediately preceding or proceeding segment) a lag of -1 or +1, respectively.  The segment following the next was assigned a lag of 2, and so on.  We tagged offscreen events using half steps.  For example, an offscreen event that occurred after the prior segment but before the just-watched segment would be assigned a lag of -0.5.

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=0.9\textwidth]{results3}
  \caption{\textbf{Characters' references drive participants' retrodiction and prediction performance.}  \textbf{A. Illustration of annotation approach.}  We manually annotated references to events in past or future segments in characters' spoken conversations.  We matched each such reference with its corresponding storyline event (and its corresponding segment number for onscreen events, or half-step segment number for offscreen events).  We then tracked the hit rate separately for referenced versus unreferenced events, in participants' uncued retrodictions and predictions.  \textbf{B. Reference rate as a function of lag.}  Across all possible just-watched segments (lag 0), the bar heights denote the average proportions of events referenced in other past (orange, negative lags) or future (blue, positive lags) segments.  \textbf{C. Difference in hit rates between all events and unreferenced events.}  To highlight the effect of characters' references to past and future events on participants' retrodictions and predictions, here we display the difference in across-segment mean hit rates between all events and unreferenced events, as a function of temporal distance (lag) to the just-watched segment.  \textbf{D. Hit rates for unreferenced events.}  The average response hit rates for unreferenced events are displayed as a function of temporal distance to the just-watched segment.  Error bars denote bootstrapped 95\% confidence intervals.  Panels B--D: colors are described in the Figure~\ref{fig:result2} caption.  \textbf{E.  Hit rates and counts of referenced and unreferenced events.}  As a function of temporal distance to the just-watched segment, the sub-panels display the across-segment mean numbers ($x$-axes) and hit rates ($y$-axes) of referenced (red) and unreferenced (gray) events that participants hit (darker shading) or missed (lighter shading) in their uncued retrodictions (top sub-panel) and uncued predictions (bottom sub-panel).}
  \label{fig:result3}
\end{figure}

@@

An additional ambiguous case arose in one participant's responses pertaining to segment 12, storyline 2, whereby the participant correctly identified an onscreen event that had not been included in our original annotations.  To account for this participant's response, we retroactively added that event to our annotations of that segment.  We also identified and counted unmatched events in participants' responses (i.e., events that did not match any annotated events).  In several cases, the two coders' independent scoring disagreed.  These cases were resolved through discussions between the two coders.

To estimate the semantic similarities between pairs of responses, we first transformed each response into a 512-dimensional vector (embedding) using \textit{Universal Sentence Encoder}~\citep[Transformer USE, ][]{CerEtal18}.  We defined \textit{similarity} as the cosine of the angle formed by the responses' vectors.  Following~\cite{HeusEtal21}, we defined the \textit{precision} of participants' responses as the median similarity between that response's vector and the embedding vectors for all other participants' recalls of the target segment.  We defined the \textit{convergence} of a given response as the mean similarity between that response's vector and all other participants' responses to the corresponding segment, in the same condition.  To compute these median or mean similarities we first applied the Fisher $z$-transformation to the similarity values, then took the median or mean of the $z$-transformed similarities, and finally applied the inverse $z$-transformation to obtain the precision or convergence score.
